---
output:
  pdf_document: default
  html_document: default
---
```{r global_options, include=FALSE}
library(knitr)
opts_chunk$set(fig.align = "center", fig.height = 5, fig.width = 6)
library(tidyverse)
theme_set(theme_bw(base_size = 12))
library(ggthemes)
```

## Project 1
*Meghana V Palukuri*

### Instructions

This knitted R Markdown document (as a PDF) *and* the raw R Markdown file (as .Rmd) should both be submitted to Canvas by 4:00pm on **Feb 26th, 2019**. These two documents will be graded jointly, so they must be consistent (as in, don't change the R Markdown file without also updating the knitted document!).

All results presented *must* have corresponding code. **Any answers/results given without the corresponding R code that generated the result will be considered absent.** To be clear: if you do calculations by hand instead of using R and then report the results from the calculations, **you will not receive credit** for those calculations. All code reported in your final project document should work properly. Please do not include any extraneous code or code which produces error messages. (Code which produces warnings is acceptable, as long as you understand what the warnings mean.)

For this project, you will be using the dataset `flavors_of_cacao`. This dataset contains expert ratings of over 1,700 individual chocolate bars, along with information on their regional origin, percentage of cocoa, the variety of chocolate bean used, and where the beans were grown.
```{r}
flavors_of_cacao <- 
  read_csv("https://raw.githubusercontent.com/clauswilke/dviz.supp/master/data-raw/cacao/cacao_clean.csv") %>%
  extract(cocoa_percent, "cocoa_percent", regex = "([^%]+)%", convert = TRUE)
head(flavors_of_cacao)
```

The column contents are as follows:

+ **company**: name of the company manufacturing the bar.
+ **bean_origin_detailed**: the specific geo-region of origin of the bar.
+ **REF**: a value linked to when the review was entered in the database. Higher = more recent.
+ **review_date**: date of publication of review.
+ **cocoa_percent**: cocoa percentage (darkness) of the chocolate bar being reviewed.
+ **location**: manufacturer base country.
+ **rating**: expert rating for the bar.
+ **bean_type**: the variety (breed) of bean used, if provided.
+ **bean_origin**: the broad geo-region of origin of the bean.

### Problems

**Problem 1: (10 pts)**
Write R code that counts the number of reviews for each company location and calculates a minimum and a maximum ratings of each company location. Filter your output for countries with more than 20 reviews, and order your output from highest to lowest number of reviews. 
```{r}

df <- flavors_of_cacao %>% 
  group_by(location) %>% 
  summarize(num_reviews = n(), min_rating = min(rating), max_rating = max(rating)) %>% 
  # Filtering for all locations with number of reviews more than 20 
  filter(num_reviews > 20) %>% 
  arrange(desc(num_reviews))
  

df

```

**Problem 2: (20 pts)** 
Use the data-frame you generated in Problem 1 to find a location with the highest maximum rating and a location with the lowest minimum ratings. Perform a statistical test to determine whether there is a significant difference in ratings between these two locations.

```{r}

# Finding the entry with the highest max rating 
df %>% 
  filter(max_rating == max(df$max_rating)) -> highest_max

# Finding the entry with the lowest min rating 
df %>% 
  filter(min_rating == min(df$min_rating)) -> lowest_min

highest_max
lowest_min
```
We can see that Italy has the highest max rating and Belgium has the lowest min rating. 
```{r}

# Finding all entries for country with highest max rating 
flavors_of_cacao %>% 
  filter(location == highest_max$location) -> highest_max_loc

# Finding all entries for country with lowest min rating 
flavors_of_cacao %>% 
  filter(location == lowest_min$location) -> lowest_min_loc

# Extracting ratings of both countries 
highest_max_loc_ratings <- highest_max_loc$rating
lowest_min_loc_ratings <- lowest_min_loc$rating

t.test(highest_max_loc_ratings,lowest_min_loc_ratings)
```

From the t-test, we can see that there is'nt a significant difference between the two locations as the p-value > 0.05 and the means are close to each other (3.325397, 3.093750). 

**Problem 3: (40 pts)**
Make one plot that visualizes the relationship between the number of reviews and maximum and minimum ratings. Use the data-frame you created in Problem 1. Your code should be well-commented and describe the various steps you take to create this figure. **HINT**: Convert your dataset to a tidy format before you plot.

**a. (30 points)**

```{r}

# Collapsing min_rating, max_rating columns into one rating column,
# with another column to indicate whether the rating is min or max
df_tidy <- df %>% 
  gather(rating_type,rating,min_rating,max_rating)

df_tidy

# Plotting using points with additional random points
ggplot(df_tidy,aes(x=rating,y=num_reviews,color=rating_type)) + geom_point() + geom_jitter()

```

**b. (10 points)** Discuss the information (overarching trends, patterns, etc.) your plot reveals. Be sure to include in your discussion the similarities/differences among minimum and maximum ratings. Your discussion should also explain the results of the t-test in Problem 2 in the context of this plot. Be sure to also include a clear, logical justification for why you selected the particular geom(s) used to represent this data. Please limit your full response to a maximum of 10 sentences.

From the plot, we can see that there are very less number of reviews for the highest max rating (5) and the lowest min rating (1), corresponding to Italy and Belgium as found in problem 2. From the plot, we can see that if we were to calculate a mean rating using all the countries, it would be around 3, corresponding to the t-test mean values for Italy and Belgium. The largest number of reviews is a clear outlier, with minumum rating around 1.5 and max rating around 4.  Minimum ratings are spread out from 1 to 3 (following a gaussian like distribution when the outlier is excluded), while max ratings are saturated around 4 (like a delta function). geom_point was used as there are multiple values of number of reviews for a particular rating value and these can be clearly visible as dots, as compared to a bar/line plot which would make the distinctions less visible. geom_jitter was used as it is a small dataset and closely generated random variables would help visualize trends better. 

**Problem 4: (30 pts)**
Think of **one** (and only one!) conceptual question to ask about the data set `flavors_of_cacao`. Clearly state your question in the space provided below. Use the ggplot2 library to create a plot that can help you find an answer to the question. For the plot, provide a clear explanation as to why this type of plot (e.g. boxplot, barplot, histogram, etc.) is best for providing the information you are asking about. Answer your question by interpreting your plot and identifying any trends it reveals, or does not reveal, as the case may be. Please limit the discussion to 4-6 sentences.

*To receive full credit for Problem 4, we look for the following for a question:*

- *A clear, coherent question about the data. (Questions end in a question mark!)*
- *The question should be conceptual and **should not** prompt a specific analysis or plot*.
- *A plot that helps answer your proposed question, with a justification for why you chose to make the type of plot that you made.*
- *An interpretation of your plot and a response to your proposed question.*
- *Statistical analysis **is not** necessary. Just interpret your plot.*

**You cannot reuse the questions about the `flavors_of_cacao` data set from the previous problems.**


**Question**

*Do major origin locations of the bean (at country level) affect ratings?*

```{r,fig.width=10,fig.height=30}
# changing size to enlarge and fit figure 
theme_set(theme_bw(base_size = 30))

# Finding major bean origin locations 
orig_locs <- flavors_of_cacao %>% 
  count(bean_origin) %>% 
  # Filtering for number of reviews greater than 10 to find major origin locations. 
  filter(n>10) %>% 
  na.omit()

df_new <- flavors_of_cacao %>% 
  filter(bean_origin %in% orig_locs$bean_origin) %>% 
  group_by(bean_origin) 

ggplot(df_new,aes(x=bean_origin,y=rating)) + geom_boxplot() + coord_flip()

```

To answer this question (of finding the relation between bean origin location and ratings), we use boxplots of the ratings for each of the locations of origin of the bean, as they help to visualize the entire range of the rating values. We can see that the location of origin of the bean does have some influence on the ratings, although countries generally have most of their ratings around the range of 3 to 3.5. From the plot, Honduras has the highest median ratings while Mexico and Ghana have the lowest median ratings. 